Tree height and hydraulic traits shape growth responses across droughts in a temperate broadleaf forest

### 2020-07-23 manuscript/response_to_reviews.Rmd

69 this is a very important point—microsite is crucial both for the ultimate height an individual grows to as well as its vulnerability. The generalization that on a global level large trees seem disproportionately vulnerable is often greeted with the skeptical observation that in specific cases a small tree died next to a large one that survived. However, the prediction is that *all else being equal* large individuals should be more vulnerable than smaller *conspecifics*. There is so much variation in microsite and other factors that the greater vulnerability of larger individuals should emerge clearly only in carefully controlled experiments where all else is indeed (virtually) equal, or with high sample sizes where variation in microsite etc should random out and the signal of generally higher vulnerability should come through.

**Thank you.**

73 The “greater root reach” point is almost universally appealed to—cites for these counterarguments would be useful.

**We now address this in the discussion: "Belowground, taller trees would tend to have larger root systems [@enquist_global_2002], but this does not necessarily imply that they have greater access to or reliance on deep soil-water resources that may be critical during drought. While tree size can correlate with the depth of water extraction [@brum_2018_hydrological], the linkage is not consistent. Shorter trees can vary broadly in the depth of water uptake [@stahl_2013_depth], and larger trees may allocate more to abundant shallow roots that are beneficial for taking up water from rainstorms [@meinzer_partitioning_1999]. Moreover, reliance on deep soil-water resources can actually prove a liability during severe and prolonged drought, as these can experience more intense water scarcity relative to non-drought conditions [@chitra-tarak_roots_2018]."**

### 2020-07-23 manuscript/climate_sensitivity_draft.Rmd
more rooting depth refs

Belowground, taller trees would tend to have larger root systems [@enquist_global_2002], but this does not necessarily imply that they have greater access to or reliance on deep soil-water resources that may be critical during drought.
Rather, larger trees may allocate more to abundant shallow roots that are beneficial for taking up water from rainstorms [@meinzer_partitioning_1999].
In any case, the potentially greater access to water did not override the disadvantage conferred by height--and, in fact, greater moisture access in non-drought years (here, higher TWI) appears to make trees more sensitive to drought [@zuleta_drought-induced_2017; @stovall_tree_2019].
This may be because moister habitats would tend to support species and individuals with more mesophytic traits [@bartlett_drought_2016; @mencuccini_ecological_2003; @medeiros_extensive_2019], potentially growing to greater heights (e.g., @detto_hydrological_2013), and these are then more vulnerable when drought hits.
The observed height-sensitivity of $Rt$, together with the lack of conferred advantage to large stature in drier topographic positions, agrees with the concept that physiological limitations to transpiration under drought shift from soil water availability to the plant-atmosphere interface as forests age [@bretfeld_plant_2018], such that tall, dominant trees are the most sensitive in mature forests. 
Again, additional research comparing drought responses across forests with different tree heights and water availability would be valuable for disentangling the relative importance of above- and belowground mechanisms across trees of different size.

The development of tree-ring chronologies for the twelve most dominant tree species at our site [@helcoski_growing_2019; @bourg_initial_2013] gave us the sample size to compare historical drought responses across species (Fig. **3**) and associated traits at a single site [see also @elliott_forest_2015]. 
Our study reinforced current understanding (see Introduction) that wood density and $LMA$ are not reliably linked to drought resistance (Table 1).
Contrary to previous studies in temperate deciduous forests, we did not find an association between xylem porosity and drought tolerance, as the two diffuse-porous species, *Liriodendron tulipifera* and *Fagus grandifolia*, were at opposite ends of the $Rt$ spectrum (Fig. **3**). 
While the low $Rt$ of *L. tulipifera* is consistent with other studies [@elliott_forest_2015], the high $Rt$ of *F. grandifolia* contrasts with studies identifying diffuse porous species in general [@elliott_forest_2015; @kannenberg_linking_2019], and the genus *Fagus* in particular [@friedrichs_species-specific_2009], as drought sensitive.
There are two potential explanations for this discrepancy.
First, other traits can and do override the influence of xylem porosity on drought resistance. 
Ring-porous species are restricted mainly to temperate deciduous forests [@wheeler_variations_2007], while highly drought-tolerant diffuse-porous species exist in other biomes (*REFS*). 
*Fagus grandifolia* had intermediate $\pi_{tlp}$ and low $PLA_{dry}$ (Fig. **S4**), which would have contributed to its drought resistance (Fig. **4**; see discussion below). 
A second explanation of why *F. grandifolia* trees at this particular site had higher $Rt$ is that the sampled individuals, reflective of the population within the plot, are generally shorter and in less-dominant canopy positions compared to most other species (Fig. **S4**).
The species, which is highly shade-tolerant, also has deep crowns [@andersonteixeira_size-related_2015], implying that a lower proportion of leaves would be affected by harsher microclimatic conditions at the top of the canopy under drought (Fig. **2**).
Thus, the high $Rt$ of the sampled *F. grandifolia* population can be explained by a combination of fairly drought-resistant leaf traits, shorter stature, and a buffered microenvironment.

>>>>

Belowground, taller trees would tend to have larger root systems [@enquist_global_2002], but this does not necessarily imply that they have greater access to or reliance on deep soil-water resources that may be critical during drought.
While tree size can correlate with the depth of water extraction [@brum_hydrological_2018], the linkage is not consistent. Shorter trees can vary broadly in the depth of water uptake [@stahl_depth_2013], and larger trees may allocate more to abundant shallow roots that are beneficial for taking up water from rainstorms [@meinzer_partitioning_1999].
Moreover, reliance on deep soil-water resources can actually prove a liability during severe and prolonged drought, as these can experience more intense water scarcity relative to non-drought conditions [@chitra-tarak_roots_2018].
In any case, the potentially greater access to water did not override the disadvantage conferred by height--and, in fact, greater moisture access in non-drought years (here, higher TWI) appears to make trees more sensitive to drought [@zuleta_drought-induced_2017; @stovall_tree_2019].
This may be because moister habitats would tend to support species and individuals with more mesophytic traits [@bartlett_drought_2016; @mencuccini_ecological_2003; @medeiros_extensive_2019], potentially growing to greater heights (e.g., @detto_hydrological_2013), and these are then more vulnerable when drought hits.
The observed height-sensitivity of $Rt$, together with the lack of conferred advantage to large stature in drier topographic positions, agrees with the concept that physiological limitations to transpiration under drought shift from soil water availability to the plant-atmosphere interface as forests age [@bretfeld_plant_2018], such that tall, dominant trees are the most sensitive in mature forests. 
Again, additional research comparing drought responses across forests with different tree heights and water availability would be valuable for disentangling the relative importance of above- and belowground mechanisms across trees of different size.

The development of tree-ring chronologies for the twelve most dominant tree species at our site [@helcoski_growing_2019; @bourg_initial_2013] gave us the sample size to compare historical drought responses across species (Fig. **3**) and associated traits at a single site [see also @elliott_forest_2015]. 
Our study reinforced current understanding (see Introduction) that wood density and $LMA$ are not reliably linked to drought resistance (Table 1).
Contrary to previous studies in temperate deciduous forests, we did not find an association between xylem porosity and drought tolerance, as the two diffuse-porous species, *Liriodendron tulipifera* and *Fagus grandifolia*, were at opposite ends of the $Rt$ spectrum (Fig. **3**). 
While the low $Rt$ of *L. tulipifera* is consistent with other studies [@elliott_forest_2015], the high $Rt$ of *F. grandifolia* contrasts with studies identifying diffuse porous species in general [@elliott_forest_2015; @kannenberg_linking_2019], and the genus *Fagus* in particular [@friedrichs_species-specific_2009], as drought sensitive.
There are two potential explanations for this discrepancy.
First, other traits can and do override the influence of xylem porosity on drought resistance. 
Ring-porous species are restricted mainly to temperate deciduous forests [@wheeler_variations_2007], while highly drought-tolerant diffuse-porous species exist in other biomes (**REFS**). 
*Fagus grandifolia* had intermediate $\pi_{tlp}$ and low $PLA_{dry}$ (Fig. **S4**), which would have contributed to its drought resistance (Fig. **4**; see discussion below). 
A second explanation of why *F. grandifolia* trees at this particular site had higher $Rt$ is that the sampled individuals, reflective of the population within the plot, are generally shorter and in less-dominant canopy positions compared to most other species (Fig. **S4**).
The species, which is highly shade-tolerant, also has deep crowns [@andersonteixeira_size-related_2015], implying that a lower proportion of leaves would be affected by harsher microclimatic conditions at the top of the canopy under drought (Fig. **2**).
Thus, the high $Rt$ of the sampled *F. grandifolia* population can be explained by a combination of fairly drought-resistant leaf traits, shorter stature, and a buffered microenvironment.
